Design and Implementation of a Multithreaded Virtual Machine for Executing Linear Logic Programs

### 2014-08-08 paper.tex
final changes.

\begin{table}[ht]
\centering
{\scriptsize
\begin{tabular}{llcl}
Program          & $Prog$ & $::=$ & $\Sigma, D$ \\
Set of Rules     & $\Sigma$ & $::=$ & $\cdot \; | \; \Sigma, R$\\
Database         & $D$ & $::=$ & $\Gamma; \Delta$ \\
Rule             & $R$ & $::=$ & $BE \lolli HE \; | \; \forall_{x}. R$ \\
Body Expression  & $BE$ & $::=$ & $L \; | \; P \; | \; C \; | \; BE, BE \; | \; \exists_{x}. BE \; | \; 1$\\
Head Expression  & $HE$ & $::=$ & $L \; | \; P \; | \; HE, HE \; | \; CE \; | \; AE \; | \; 1$\\
Linear Fact      & $L$ & $::=$ & $l(\hat{x})$\\
Persistent Fact  & $P$ & $::=$ & $\bang p(\hat{x})$\\
Constraint       & $C$ & $::=$ & $c(\hat{x})$ \\
Comprehension    & $CE$ & $::=$ & $\comprehension$ \\
Aggregate        & $AE$ & $::=$ & $\aggregate$ \\
Operations       & $A$ & $::=$ & $\mathtt{min} \; | \; \mathtt{max} \; | \; \mathtt{sum} \; | \; \mathtt{count}$ \\
Sub-Head         & $SH$ & $::=$ & $L \; | \; P \; | \; SH, SH \; | \; 1$\\
Linear Facts     & $\Delta$ & $::=$ & $\cdot \; | \; \Delta, l(\hat{t})$ \\
Persistent Facts & $\Gamma$ & $::=$ & $\cdot \; | \; \Gamma, \bang p(\hat{t})$ \\
\end{tabular}
}
\caption{Abstract syntax of LM}
\label{tbl:ast}
\end{table}

@@

During execution, threads can steal nodes of other threads to keep
themselves busy. The load balancing aspect of the system is performed
by our work scheduler that is based on a simple work stealing
algorithm. The pseudo-code for the main thread loop is shown in
Fig.~\ref{code:work_loop}. In each round, a thread inspects its work
queue for active nodes with new candidate rules and, if there is any,
procedure \texttt{process\_node()} executes it. If the work queue is
empty, the thread attempts to steal one node from another
thread. Starting from a random thread, it cycles through all the
threads to find one active node. Eventually, there will be no more
work to do and the threads will go idle. There is a global atomic
counter, a global boolean flag and one boolean flag for each thread
that are used to detect termination. Once a thread goes idle, it
decrements the global counter and changes its flag to idle. If the
counter reaches zero, the global flag is set to idle. Since every
thread will be busy-waiting and checking the global flag, they will
detect the change and stop executing.

@@

Figure~\ref{fig:overview} presents the layout of our virtual machine
for a program with six nodes and two running threads. Each thread
space includes the nodes owned by the thread (the dotted arrows
represent the edges between nodes) and a \emph{Work Queue}, a single
linked list containing \emph{active nodes}, i.e., nodes that have new
facts to process. Initially, the \emph{Work Queue} is filled with all
the nodes of the thread in order to derive the
axioms. Figure~\ref{fig:overview} also illustrates the internal
structure layout of a node, which includes: the database of linear
facts (\emph{Linear DB}); the database of persistent facts
(\emph{Persistent DB}); the rule matching structures (\emph{Rule
Engine}); and an auxiliary buffer for storing intermediate facts
coming from other threads (\emph{Fact Buffer}).

@@

Several linear logic programming languages have been developed in the past~\cite{Miller85anoverview}.
Lolli, a programming language based on a fragment of intuitionistic linear logic~\cite{Hodas94logicprogramming}, 
proves goals by lazily managing the context of linear resources during backward-chaining proof search.
LolliMon~\cite{Lopez:2005:MCL:1069774.1069778} is a concurrent linear logic programming language that integrates
both forward and backward-chaining search, where the backward-chaining phase is done sequentially but
the forward-chaining is done concurrently inside a monad. The backward-chaining phase is suspended
between the forward-chaining phases, where a fix-point is computed.
LolliMon is derived from the concurrent logical framework called CLF~\cite{Watkins:2004uq}. 
\makeatletter{}\section{Conclusions}

@@

@@

\bibitem[L\'{o}pez et~al.(2005)L\'{o}pez, Pfenning, Polakow, and
  Watkins]{Lopez:2005:MCL:1069774.1069778}
P.~L\'{o}pez, F.~Pfenning, J.~Polakow, and K.~Watkins.
\newblock Monadic concurrent linear logic programming.
\newblock In \emph{Proceedings of the 7th ACM SIGPLAN International Conference
  on Principles and Practice of Declarative Programming}, PPDP '05, pages
  35--46, New York, NY, USA, 2005.

@@

>>>>

\begin{table}[ht]
\centering
{\scriptsize
\begin{tabular}{llcl}
Program          & $Prog$ & $::=$ & $\Sigma, D$ \\
Set of Rules     & $\Sigma$ & $::=$ & $\cdot \; | \; \Sigma, R$\\
Database         & $D$ & $::=$ & $\Gamma; \Delta$ \\
Rule             & $R$ & $::=$ & $BE \lolli HE \; | \; \forall_{x}. R$ \\
Body Expression  & $BE$ & $::=$ & $L \; | \; P \; | \; C \; | \; BE, BE \; | \; \exists_{x}. BE \; | \; 1$\\
Head Expression  & $HE$ & $::=$ & $L \; | \; P \; | \; HE, HE \; | \; CE \; | \; AE \; | \; 1$\\
Linear Fact      & $L$ & $::=$ & $l(\hat{x})$\\
Persistent Fact  & $P$ & $::=$ & $\bang p(\hat{x})$\\
Constraint       & $C$ & $::=$ & $c(\hat{x})$ \\
Comprehension    & $CE$ & $::=$ & $\comprehension$ \\
Aggregate        & $AE$ & $::=$ & $\aggregate$ \\
Operations       & $A$ & $::=$ & $\mathtt{min} \; | \; \mathtt{max} \; | \; \mathtt{sum} \; | \; \mathtt{count}$ \\
Sub-Head         & $SH$ & $::=$ & $L \; | \; P \; | \; SH, SH \; | \; 1$\\
Linear Facts     & $\Delta$ & $::=$ & $\cdot \; | \; \Delta, l(\hat{t})$ \\
Persistent Facts & $\Gamma$ & $::=$ & $\cdot \; | \; \Gamma, \bang p(\hat{t})$ \\
Terms            & $t$ & $::=$ & $\m{node}(n) \; | \; \m{int}(n) \; | \; \m{float}(f) \; | $\\
                 &     &       & $ \m{string}(s) \; | \; \m{bool}(b) \; | \; l \; | \; \m{pair}(t,t)$ \\
Lists            & $l$ & $::=$ & $\m{nil} \; | \; [ t \; | \; l ]$\\
\end{tabular}
}
\caption{Abstract syntax of LM}
\label{tbl:ast}
\end{table}

@@

Figure~\ref{fig:overview} presents the layout of our virtual machine
for a program with six nodes and two running threads. Each thread
space includes the nodes owned by the thread (the dotted arrows
represent the edges between nodes) and a \emph{Work Queue}, a single
linked list containing \emph{active nodes}, i.e., nodes that have new
facts to process. Initially, the \emph{Work Queue} is filled with all
the nodes of the thread in order to derive the axioms.

During execution, threads can steal nodes of other threads to keep
themselves busy. The load balancing aspect of the system is performed
by our work scheduler that is based on a simple work stealing
algorithm. The pseudo-code for the main thread loop is shown in
Fig.~\ref{code:work_loop}. In each round, a thread inspects its \emph{Work Queue}
for active nodes with new candidate rules and, if there is any,
procedure \texttt{process\_node()} is called on the target node.
Otherwise, if the \emph{Work Queue} is empty, the thread attempts to
steal one node from another
thread. Starting from a random thread, it cycles through all the
threads to find one active node. Eventually, there will be no more
work to do and the threads will go idle. There is a global atomic
counter, a global boolean flag and one boolean flag for each thread
that are used to detect termination. Once a thread goes idle, it
decrements the global counter and changes its flag to idle. If the
counter reaches zero, the global flag is set to idle. Since every
thread will be busy-waiting and checking the global flag, they will
detect the change and stop executing.

@@

Figure~\ref{fig:overview} also illustrates the internal
structure layout of a node, which includes: the database of linear
facts (\emph{Linear DB}); the database of persistent facts
(\emph{Persistent DB}); the rule matching structures (\emph{Rule
Engine}); and an auxiliary buffer for storing intermediate facts
coming from other threads (\emph{Fact Buffer}).

@@

Several linear logic programming languages have been developed in the past~\cite{Miller85anoverview}.
Lolli, a programming language based on a fragment of intuitionistic linear logic~\cite{Hodas94logicprogramming}, 
proves goals by lazily managing the context of linear resources during backward-chaining proof search.
Forum~\cite{Miller94} is an extension of Lolli that is based on a classical proof system. While Lolli uses an intuitionistic proof system that supports only one conclusion, Forum allows multiple conclusions. Forum is therefore better adapted to represent concurrent computations than Lolli, since many multiple goals need to be proved at once.

Another extension of Lolli is LolliMon~\cite{Lopez:2005:MCL:1069774.1069778}, a concurrent linear logic programming language that integrates
both forward and backward-chaining search, where the backward-chaining phase is done sequentially but
the forward-chaining is done concurrently inside a monad. The backward-chaining phase is suspended
between the forward-chaining phases, where a fix-point is computed.
LolliMon is derived from the concurrent logical framework called CLF~\cite{Watkins:2004uq}.

Lygon~\cite{Harland96} is a backward-chaining linear logic programming language that extends Prolog with linear resources. Lygon is not only able to run Prolog programs but also allows the body of clauses to be linear that can be used exactly once in each query.
In contrast to Prolog, where there is no clause selection, Lygon needs to use heuristics to decide which clause to prove first, since that will help finding a proof and finding it efficiently. Notably, Lygon is very suitable to solve graph-based problems due to the mutable state provided by linear clauses.

\makeatletter{}\section{Conclusions}

@@

\bibitem[Harland(1996)]{Harland96}
J.~Harland, D.~Pym and M.~Winikoff
\newblock Programming in Lygon: An Overview
\newblock In \emph{International Conference on Algebraic Methodology and Software Technology (AMAST)}, 391-405,  1996.

@@

\bibitem[L\'{o}pez et~al.(2005)L\'{o}pez, Pfenning, Polakow, and
  Watkins]{Lopez:2005:MCL:1069774.1069778}
P.~L\'{o}pez, F.~Pfenning, J.~Polakow, and K.~Watkins.
\newblock Monadic concurrent linear logic programming.
\newblock In \emph{International Conference on Principles and Practice of Declarative Programming}, PPDP '05, pages
  35--46, New York, NY, USA, 2005.

@@

\bibitem[Miller(1994)]{Miller94}
D.~Miller.
\newblock A multiple-conclusion meta logic
\newblock In \emph{Logic in Computer Science (LICS)}, pages 272--281, 1994.
